Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-10 submission/manuscript.rmd
first draft of GUSTO results

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
rmse_scenarios_10_12_16 <- readr::read_csv(
  file = here::here(file.path("data/processed", "rmse.csv"))
) %>%
  dplyr::filter(scenarioId %in% c(10, 12, 16))
```
Under the base case of constant relative treatment effect, the model assuming a
constant treatment effect had the lowest median RMSE, regardless of true
prediction AUC and sample size. Linear interaction models demonstrated
comparable performance. Among the RCS smoothing methods, the one fitted with 3
knots always performed best, while the increased flexibility achieved when
increasing the knots resulted in overfitting and worse performance. The adaptive
approach under all scenarios performed similar to the model with smaller RMSE in
all scenarios.

@@

@@

When we introduced deviations from the base case of constant relative treatment
effects, while keeping fixed both the sample size (N = 4250) and the true
prediction AUC (0.75) the linear interaction model had the lowest RMSE (Figure
\@ref(fig:rmsebase); Panels A, D, and G). When these deviations were moderate
(Figure \@ref(fig:rmsebase); Panel A) the constant treatment effect model had
comparable performance to the linear interaction model. This can be attributed
to the fact that such deviations are quite mild and absolute benefits maintain
similar patterns across baseline risk. On the contrary, when strong quadratic
deviations were considered the constant effect model's RMSE sharply increased,
while the more flexible method of RCS smoothing (3 knots) preformed very well
(Figure \@ref(fig:rmsebase); Panel G). Again, increasing the number of knots
increased RMSE, indicating overfitting.

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE when strong quadratic deviations from the base
case of constant relative treatment effects were assumed (Figure
\@ref(fig:rmsebase); Panel H). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \@ref(fig:rmsebase); Panels B and E). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations. We observed similar
results when we increased the sample size to 17000, while keeping the true
prediction AUC constant at 0.75 (Figure \@ref(fig:rmsebase); Panels C, F, and I).

```{r, echo=FALSE, warning=FALSE, message=FALSE}
rmse_scenarios_10_64_66_65 <- readr::read_csv(
  file = here::here(file.path("data/processed", "rmse.csv"))
) %>%
  dplyr::filter(scenarioId %in% c(10, 64, 66, 65))
```

When focusing on the different scenarios where true treatment-covariate
interactions were considered all methods had similar RMSE performance. In case
of strong treatment-covariate interactions the constant effect model had
slightly increased RMSE
(`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(constant_treatment_effect) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`])
compared to the other methods. The linear interaction model with the risk linear
predictor had the lowest RMSE
(`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% median() %>% round(3)`; [`r rmse_scenarios_10_64_66_65 %>% filter(scenarioId == 65) %>% select(linear_predictor) %>% unlist() %>% quantile(c(.025, .975)) %>% round(3) %>% paste(collapse = ", ")`]).

@@

## Real data
```{r gusto, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/gusto.png"))
```

>>>>

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(SmoothHte)
library(rms)

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios where the base case of true constant relative treatment effect
(OR = 0.8, N = 4250 and AUC = 0.75) or moderate relative deviations were
considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
strong linear and quadratic deviations from the base case the linear interaction
model performed best (Figure \@ref(fig:rmsebase); Panels B and C). Only in the
case of strong quadratic deviations models including RCS smoothing (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS smoothing resulted in higher error rates across all scenarios.
The performance of the adaptive approach was quite comparable with the best
performing model in all scenarios.

@@

When we increased the sample size (N = 17000), in all scenarios the model more
in agreement with the underlying settings had the lowest error rates. Under the
base case of constant relative treatment effects (OR = 0.8), the model including
a constant relative treatment effect had the lowest RMSE (Figure
\@ref(fig:rmsesamplesize); Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\@ref(fig:rmsesamplesize); Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \@ref(fig:rmsesamplesize); Panel D).

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE when strong quadratic deviations from the base
case of true constant relative treatment effects were assumed (Figure
\@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \@ref(fig:rmseauc); Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations.

@@

@@

## Real data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
load(here::here("data/raw/gusto.rda"))
gusto <- gusto %>%
  tibble() %>%
  filter(!is.na(tpa))

treatmentArms <- gusto %>%
  group_by(tpa) %>%
  summarise(n = n())
```

We demonstrate the different methods for individualizing treatment benefits
using data from `r nrow(gusto)` patients with an acute myocardial infarction
(MI) included in the GUSTO-I trial.
`r treatmentArms %>% filter(tpa == 1) %>% select(n)` patients were randomized to
tissue plasminogen activator (tPA) treatment and
`r treatmentArms %>% filter(tpa == 0) %>% select(n)` were randomized to
streptokinase. The outcome of interest was 30-day mortality, recorded for all
patients.

In line with (Califf 1997; Steyerberg 2000), we fitted a logistic regression
model with 6 baseline covariates, i.e. age, Killip class, systolic blood
pressure, heart rate, an indicator of previous MI, and the location of MI, to
predict 30-day mortality risk. A constant effect of treatment was included in
the model. When deriving risk predictions for individuals we set the treatment
indicator to 0. More information on model development can be found in the
supplement.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
gusto <- gusto %>%
  tibble() %>%
  rename(
    "outcome" = "day30",
    "treatment" = "tpa"
  )

prediction <- lrm(
  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
  data = gusto,
  maxit = 99
)

riskLinearPredictor <- predict(
  prediction,
  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
)

gusto <- gusto %>%
  mutate(
    riskLinearPredictor = riskLinearPredictor
  )

constantModel <- fitModelBasedHte(
  data     = gusto, 
  settings = createModelBasedSettings(type = "treatment")
)

tmp <- gusto %>%
  mutate(
    predictedBenefit = predictBenefitModelBasedHte(
      p             = plogis(.$riskLinearPredictor), 
      modelBasedFit = constantModel
    )
  )

constantC   <- calculateCForBenefit(tmp)
constantIci <- calculateCalibrationForBenefit(tmp)

stratified <- fitStratifiedHte(
  data     = gusto,
  settings = createStratifiedSettings()
)

linearModel <- fitModelBasedHte(
  data     = gusto,
  settings = createModelBasedSettings()
)

tmp <- gusto %>%
  mutate(
    predictedBenefit = predictBenefitModelBasedHte(
      p             = plogis(.$riskLinearPredictor), 
      modelBasedFit = linearModel
    )
  )

linearC   <- calculateCForBenefit(tmp)
linearIci <- calculateCalibrationForBenefit(tmp)

rcs3Model <- fitRcsHte(
  data     = gusto,
  settings = createRcsSettings()
)

tmp <- gusto %>%
  mutate(
    predictedBenefit = predictSmoothBenefit(
      p          = plogis(riskLinearPredictor), 
      smoothFit  = rcs3Model
    )
  )

rcs3C   <- calculateCForBenefit(tmp)
rcs3Ici <- calculateCalibrationForBenefit(tmp)

rcs4Model <- fitRcsHte(
  data     = gusto,
  settings = createRcsSettings(nKnots = 4)
)

tmp <- gusto %>%
  mutate(
    predictedBenefit = predictSmoothBenefit(
      p          = plogis(riskLinearPredictor), 
      smoothFit  = rcs4Model
    )
  )

rcs4C   <- calculateCForBenefit(tmp)
rcs4Ici <- calculateCalibrationForBenefit(tmp)

rcs5Model <- fitRcsHte(
  data     = gusto,
  settings = createRcsSettings(nKnots = 5)
)

tmp <- gusto %>%
  mutate(
    predictedBenefit = predictSmoothBenefit(
      p          = plogis(.$riskLinearPredictor), 
      smoothFit  = rcs5Model
    )
  )

rcs5C   <- calculateCForBenefit(tmp)
rcs5Ici <- calculateCalibrationForBenefit(tmp)

gustoPerformance <- tibble(
  discrimination = c(
    constantC, linearC, rcs3C,
    rcs4C, rcs5C
  ),
  calibration = c(
    constantIci$ici, linearIci$ici, rcs3Ici$ici,
    rcs4Ici$ici, rcs5Ici$ici
  ),
  method = c(
    "constant treatment effect",
    "linear interaction",
    "RCS smoothing with 3 knots",
    "RCS smoothing with 4 knots",
    "RCS smoothing with 5 knots"
  )
)

cBenefit <- gustoPerformance %>% arrange(discrimination)

```

We used the risk linear predictor to fit the the proposed methods under study
for individualizing absolute benefit predictions. All methods had quite
comparable results, in the sense that we predicted increasing benefits for
patients with higher baseline risk predictions. In terms of c-for benefit,
validated internally, all models had quite comparable performance ranging from
0.519 (RCS smoothing with 3 knots) to 0.542 (RCS smoothing with 4 knots).
Similar conclusions could be drawn in terms of ICI-for-benefit which ranged
from 0.0039 (linear interaction approach) to 0.0053 (RCS smoothing with 4
knots).

The adaptive approach picked the
model with RCS smoothing with 4 knots, which had quite comparable performance to
the smooth fit with 5 knots. However, for lower baseline risk these 2 models
predicted implausible benefits and maybe should be avoided when applying such
models in practice. The linear interaction model and the model with RCS
smoothing (3 knots) made very similar predictions, while also followed quite
closely the evolution of the stratified estimates. In this case, we observed
that even using a simple model with a constant relative treatment effect can
result in very similar benefit predictions as other more complex approaches.

```{r gusto, echo=FALSE, fig.cap="Caption", fig.show="hold", out.width = '100%'}
knitr::include_graphics(here::here("figures/gusto.png"))
```
